### 2017-12-15 Writing/disparity_review.tex
first draft of the data part

\section{Introduction}

Biodiversity is not a smooth gradient of forms, functions, ecologies, behaviours, etc., but it is rather variable and discontinuous.
This is at the base of biological sciences and is linked to the concept of disparity \citep{Wills2001,Hopkins2017} with the observation that some groups of species are more similar or dissimilar than others.
This diversity in morphologies is in fact proper in nature and decoupled from taxonomic diversity: some groups may exhibit low diversity of species by high diversity of shapes \citep[or the other way around]{ruta2013,hopkinsdecoupling2013}.
In palaeobiology, this concept, has been studied since the 90s describing \citep{gould1989wonderful,gould1991disparity,briggs1992morphological,Wills1994,Foote01071994,Foote29111996,jernvall1996molar,foote1997evolution} and as been extended with modifications to macroevolutionary analysis in general since the 2000s (whether based on continuous traits; e.g. \citealt{Harmon961,geiger2008}; or on geometric morphometrics; e.g. \citealt{claude2008morphometrics,zelditch2012geometric,adams2013geomorph,adams2017geometric}).

@@

\cite{prentice2011} define disparity as: ``a term widely (albeit not always consistently) used to describe the range of forms in a group of organisms, or the difference among different body plans''.

Are all these methods the right tools to answer these questions (described in the intro) or are they meer “inductive fishing trips” and are actually not testing any of the hypothesis related to the disparity questions:

Can the data used in disparity studies actually test the phenomenon researchers are trying to capture?
E.g. for discrete morphological characters: disparity of fossilizable characters.
E.g. for morphometrics: disparity of the skull procrustes
E.g. for dtt: disparity of body/limb/beak length
Are  these data types enough to test disparity hypothesis or do we need more data/tone down the hypothesis?

Does the transformation of this data reflect our hypothesis? E.g. a morphospace as a pairwise matrix reflects the dissimilarity between the data, a morphospace based on an ordination reflects the variability within the data, a morphospace on both reflects the variability within the dissimilarity, are each method reflecting that?

Do the disparity metrics really reflect the variation in the hypothesis that are tested?
E.g. Are the variance/range based metrics a description of the ordination or the observed biological phenomenon (i.e. they are not including covariance and phylogenetic signal)
E.g. What does the average distance between species or from a point really convey? Is the distribution not more interesting?
Do we need more like metrics measuring explicitly what is tested in the hypothesis (rather than approximate it). For example, for competitive replacement, should we not measure whether group A occupies the same place as group B through time and whether that is due to competition or simply because any other place in the space cannot be occupied (and why are such spaces unoccupied)?

Are the methods used to test hypothesis in disparity actually testing what we think they’re testing (problem of multidimensional hypothesis)
What does an anova or related variance based test real infirms/confirms regarding the disparity questions?
How do we deal with bootstrapped data and pseudo-replications?
Can we realistically use univariate gaussian models to test hypothesis based on multivariate, probably non-gaussian data?

Conclusion:
We need a more hypothesis driven way of analysing disparity. Disparity methods should be the tool for answering disparity hypothesis, not merely a description of the data.

% \section{What \textit{is} disparity (really)?}

% Disparity can describe either the metric (disparity index \citep{Hopkins2017} or disparity metric \citep{Wills2001}) or the whole pipeline \citep{Claddis,zelditch2012geometric}

% % §Methods for data:
% \subsection{Data collection methodology}
% To assess the usage of disparity in different published studies, we collected methodological data from the 500 first Google Scholar results for the key words ``morphological disparity'' per order of appearance (accessed on the 1st of November 2017).
% For the 230 relevant papers among the 500 matches, we collected the following methodological data:

% \begin{itemize}
%     \item What was the focal biological group?
%     \item What kind of data was measured (e.g. landmarks, discrete data, etc.)?
%     \item Was data collected on the full organism or not?
%     \item How was the morphospace explicitly defined (e.g. PCA, PCO, MDS, etc.)?
%     \item How was the disparity metric(s) explicitly defined?
%     \item Which statistical test was applied to test the disparity related hypothesis?
%     \item Was phylogeny taken into account or not?
% \end{itemize}

% We used only the explicit definition of the morphospace and the disparity metric(s) in this search since a few number of papers had a vague definition of either or both (e.g. a disparity metric was measured but not described anywhere in the paper).

% The remaining 270 matches were disparity was mentioned but not measured felt in the following categories: papers out of topic, papers mentioning morphological disparity without measuring it, review papers, papers not accessible (either through a broken link or a paywall) or referenced citation without the paper (as a Google Scholar match).

% To reduce the amount of categories for the 230 recorded methods, we concatenated different methods in a smaller number of categories (see supplementary materials and \url{https://github.com/TGuillerme/Disparity_Working_Group/blob/master/Analysis/data_cleaning.Rmd}).

% \begin{figure}[!htbp]
% \centering
%    \includegraphics[width=1\textwidth]{Figures/MethodsProportions.pdf} 
% \caption{Disparity methods proportional usage: (1) Data type: which input data was used (...); (2) Morphospace: how was the morphospace obtained (...); (3) disparity: what type of disparity metric was calculated (...); (4) test: what type of test was applied (...).}
% \label{Fig:MethodsProportions}
% \end{figure}

@@

% \subsection{The three main disparity analysis}
% From the collected data, we can highlight the use of three main different disparity analysis with their associated data/morphospace/metric/tests and related to specific methodological implementations.

% \begin{itemize}
%     \item \textbf{The ``\texttt{Claddis}'' approach:} this group of methods uses discrete morphological data (sometimes referred to as ``Cladistic'' characters) for the full organisms to build a PCO from the organism's pairwise distance as a morphospace. Disparity is then often measured as a variation of the ordinated matrix dimensions' variances or ranges (e.g. the sum of variance or/and the sum of ranges). Hypothesis are often tested using multivariate ANOVAs on the pairwise distance matrix or by simply comparing the confidence intervals overlap of the disparity from different groups.

%     \item \textbf{The ``\texttt{geomorph}'' approach:} this group of method is based on landmark data (2D or 3D) on parts of the organism studied usually the skull) and use a Procrustes transformation of the landmarks that are then directly ordinated using a PCA (but sometimes RWA). Disparity is often measured as a distance metric (e.g. the distance between the species and a point in the morphospace such as its centroid). Hypothesis are then tested using ANOVA type tests with usually no phylogenetic correction (although phylogeny is sometimes used to correct the morphospace).

%     \item \textbf{The ``\texttt{dtt}'' approach:} this method can directly use continuous or discrete data for the full organism without any ordination (but not necessarily), and will measure disparity as the average pairwise distance between species (whether euclidean or any other type of distance). Hypothesis of higher/lower disparity can then be measured using null evolutionary models.

% \end{itemize}

>>>>

\section{Introduction}
\label{text:intro}

Biodiversity is not a smooth gradient of forms, functions, ecologies, behaviours, etc., but it is rather variable and discontinuous. %TG: Actually is it?
This is at the base of biological sciences and is linked to the concept of disparity \citep{Wills2001,Hopkins2017} with the observation that some groups of species are more similar or dissimilar than others.
This diversity in morphologies is in fact proper in nature and decoupled from taxonomic diversity: some groups may exhibit low diversity of species by high diversity of shapes \citep[or the other way around]{ruta2013,hopkinsdecoupling2013}.
In palaeobiology, this concept, has been studied since the 90s describing \citep{gould1989wonderful,gould1991disparity,briggs1992morphological,Wills1994,Foote01071994,Foote29111996,jernvall1996molar,foote1997evolution} and as been extended with modifications to macroevolutionary analysis in general since the 2000s (whether based on continuous traits; e.g. \citealt{Harmon961,geiger2008}; or on geometric morphometrics; e.g. \citealt{claude2008morphometrics,zelditch2012geometric,adams2013geomorph,adams2017geometric}).

@@

\cite{prentice2011} define disparity as: ``a term widely (albeit not always consistently) used to describe the range of forms in a group of organisms, or the difference among different body plans''.
Disparity can describe either the metric (\citealt{Wills2001}; or index \citealt{Hopkins2017}) or the whole pipeline \citep{zelditch2012geometric,lloyd2016estimating}.
To assess the usage of disparity in different published studies, we collected methodological data from the 500 first Google Scholar results for the key words ``morphological disparity'' per order of appearance (accessed on the 1st of November 2017).
For the 230 relevant papers among the 500 matches, we collected the following methodological data: (1) What was the focal biological group? ; (2) What kind of data was measured (e.g. landmarks, discrete data, etc.)?  ; (3) Was data collected on the full organism or not? ; (4) How was the morphospace explicitly defined (e.g. PCA, PCO, MDS, etc.)? ; (5) How was the disparity metric(s) explicitly defined? ; (6) Which statistical test was applied to test the disparity related hypothesis? ; (7) Was phylogeny taken into account or not?
We used only the explicit definition of the morphospace and the disparity metric(s) in this search since a few number of papers had a vague definition of either or both (e.g. a disparity metric was measured but not described anywhere in the paper).
The remaining 270 matches were disparity was mentioned but not measured felt in the following categories: papers out of topic, papers mentioning morphological disparity without measuring it, review papers, papers not accessible (either through a broken link or a paywall) or referenced citation without the paper (as a Google Scholar match).
To reduce the amount of categories for the 230 recorded methods, we concatenated different methods in a smaller number of categories (see supplementary materials and \url{https://github.com/TGuillerme/Disparity_Working_Group/blob/master/Analysis/data_cleaning.Rmd}).

\begin{figure}[!htbp]
\centering
   \includegraphics[width=1\textwidth]{Figures/MethodsProportions.pdf} 
\caption{Disparity methods proportional usage: (1) Data type: which input data was used (...); (2) Morphospace: how was the morphospace obtained (...); (3) disparity: what type of disparity metric was calculated (...); (4) test: what type of test was applied (...).}
\label{Fig:MethodsProportions}
\end{figure}

From the collected data, we can highlight the use of three main different disparity analysis with their associated data/morphospace/metric/tests and related to specific methodological implementations.

\begin{itemize}
    \item \textbf{The ``\texttt{Claddis}'' approach:} this group of methods uses discrete morphological data (sometimes referred to as ``Cladistic'' characters) for the full organisms to build a PCO from the organism's pairwise distance as a morphospace. Disparity is then often measured as a variation of the ordinated matrix dimensions' variances or ranges (e.g. the sum of variance or/and the sum of ranges). Hypothesis are often tested using multivariate ANOVAs on the pairwise distance matrix or by simply comparing the confidence intervals overlap of the disparity from different groups.

    \item \textbf{The ``\texttt{geomorph}'' approach:} this group of method is based on landmark data (2D or 3D) on parts of the organism studied usually the skull) and use a Procrustes transformation of the landmarks that are then directly ordinated using a PCA (but sometimes RWA). Disparity is often measured as a distance metric (e.g. the distance between the species and a point in the morphospace such as its centroid). Hypothesis are then tested using ANOVA type tests with usually no phylogenetic correction (although phylogeny is sometimes used to correct the morphospace).

    \item \textbf{The ``\texttt{approach}'' approach:} this method can directly use continuous or discrete data for the full organism without any ordination (but not necessarily), and will measure disparity as the average pairwise distance between species (whether euclidean or any other type of distance). Hypothesis of higher/lower disparity can then be measured using null evolutionary models.
\end{itemize}

%
The advent of these three approaches coincides with the explosion of disparity analysis (Fig. \ref{Fig:GoogleOccurences}) and suggest a high popularity of disparity analysis.
%
Unfortunately, however, few actual tests have been performed on whether any of these approaches \textit{actually} allows to tackle the classic disparity questions (see \ref{text:intro}{Introduction}).
Are all these methods the right tools to answer these questions are they mere ``inductive fishing trips'' used to describe the data rather than testing any of the hypothesis related to the disparity questions?

\section{Does data used in disparity analysis allows to test hypothesis?}
In fact, the question of which data to analyse \citep{hetherington2015cladistic} and which part of the data to use has been analysed \citep{hopkins2017well} has rarely been assessed.
Can the data used in disparity studies actually test the phenomenon researchers are trying to capture?

For example, when studying discrete morphological characters (often following the ``\texttt{Claddis}'' approach), the data used is often recycled data used for phylogenetic analysis.
These datasets are usually design to recover evolutionary history and differentiate clades and, often due to the preservations of the fossil record, are mainly based on hard tissue characters.
In phylogenetics, this artefact have been shown to be leading to measurable topological recovery error (for soft \textit{vs} hard characters; \citealt{sansomfossilization2013}; or dental \textit{vs} cranial characters \citealt{sansom2017dental}).
Therefore, is the morphospace composed of ``fossilisable'' characters enough for answering evolutionary hypothesis?
For example, are niche replacement hypothesis equivalent to fossilisable-character-based-niches replacement hypothesis?

Similarly, many studies in geometric morphometric are based solely on skull shape using landmark techniques combined with Procrustes sumperimposition \citep[the ``\texttt{geomorph}'' approach;][]{zelditch2012geometric,adams2017geometric}.
However, it has been shown that the whole skeleton displays different integrated modules with different specific evolutionary constraints leading to variable rates and modes of evolution (\citealt{Goswami20130254}; and this even at the cranial level \citealt{goswami2010influence}).
Is skull variation enough for measuring body plans variation through time for example?

In the case where disparity is measured a single continuous trait (or a collection of traits; following the ``\texttt{geiger}'' approach), for example in \cite{price2014niche}, beak and limb length as well as body mass are used to differentiate groups of birds and different zones in the morphospace can then be interpreted as different niches.
Is this set of traits studied in isolation sufficient to test niche occupancy hypothesis?

Additionally, in many cases, the disparity metrics and hypothesis are often based on a transformation of the data (pairwise distances, PCA, PCO, MDS, RWA, etc.).
Can these transformation readily reflect the raw data in our hypothesis?
For examples a morphospace as a pairwise matrix reflects the dissimilarity between the data, a morphospace based on an ordination reflects the variability within the data, a morphospace based on both reflects the variability within the dissimilarity, are each method reflecting that?

\section{Do the disparity metrics (or indices) really reflect variation in the studied phenomenon?}
Do the disparity metrics really reflect the variation in the hypothesis that are tested?
E.g. Are the variance/range based metrics a description of the ordination or the observed biological phenomenon (i.e. they are not including covariance and phylogenetic signal)
E.g. What does the average distance between species or from a point really convey? Is the distribution not more interesting?
Do we need more like metrics measuring explicitly what is tested in the hypothesis (rather than approximate it). For example, for competitive replacement, should we not measure whether group A occupies the same place as group B through time and whether that is due to competition or simply because any other place in the space cannot be occupied (and why are such spaces unoccupied)?

\section{Is the current statistical toolkit sufficient for testing disparity hypothesis?}
Are the methods used to test hypothesis in disparity actually testing what we think they’re testing (problem of multidimensional hypothesis)
What does an anova or related variance based test real infirms/confirms regarding the disparity questions?
How do we deal with bootstrapped data and pseudo-replications?
Can we realistically use univariate gaussian models to test hypothesis based on multivariate, probably non-gaussian data?

\section{What are we missing?}

\subsection{Learning lessons from other fields}
In ecology, disparity bears strong parallels with $\beta$-diversity in ecology (a measure of ecological communities (dis)similarity): one biological observation described by a vast array of metrics \citep{baselga2010partitioning, anderson2011navigating, donohue2016navigating}.

\section{Conclusion}
We need a more hypothesis driven way of analysing disparity.
Disparity methods should be the tool for answering disparity hypothesis, not merely a description of the data.

@@
